---
title: "Calibrated inference under DP projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrated inference under DP projection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(synthexpln)
set.seed(1)
```

## 1. Why naive standard errors are too small

When an analyst releases differentially private synthetic data, whoever refits the
model on it recovers the published DP coefficients. The standard errors
they compute from the synthetic data, however, are **too small**.

The reason is that those standard errors capture only the sampling variability of
the synthetic fit. They know nothing about the DP noise that produced
$\beta_{DP} = \hat\beta + \eta_{DP}$ in the first place, because by the time the
synthetic data exists that noise has already been folded into the point estimate
and looks like signal.

The two sources are independent, so they add in variance:

$$SE_{\mathrm{corr},j} = \sqrt{SE_{\mathrm{syn},j}^2 + \sigma_{\beta,j}^2}$$

where $\sigma_{\beta,j}$ is the standard deviation of the DP noise **on the raw
scale of coefficient $j$**.

> **There is no column scale in that formula.** A divisor $s_j$ belongs there only
> when the mechanism perturbed the *standardised* coefficients, since noise of sd
> $\sigma_\beta$ on $\beta_j s_j$ lands as sd $\sigma_\beta / s_j$ on $\beta_j$.
> The sound routes do not do that: they add isotropic noise to the raw
> $\hat\beta$ at a global sensitivity. Dividing anyway shrinks the DP variance
> component by the column scale, which on a wide-scale covariate can be a factor
> of several hundred, and the "corrected" interval is barely wider than
> the uncorrected one. That understates the uncertainty of a private release,
> which is the only direction this error must not take.

Every DP generator therefore reports `$sigma.beta.raw`: a **named** vector of
raw-scale noise sds, one per coefficient. Use it, and let the package refuse
rather than guess if it cannot name them all.

## 2. Computing corrected standard errors

We use `gen.syn.dp.ols.public()`, the exact public-design route. `B.y` is a public
bound on the response, taken from the DGP on the line above and not from the
sample.

```{r}
n <- 500
d <- data.frame(x = rnorm(n))
d$y <- 1 + 0.7 * d$x + rnorm(n)

syn <- gen.syn.dp.ols.public(d, y ~ x, epsilon = 10, delta = 1e-6,
                             B.y = 8, seed = 1)

fit.syn <- lm(y ~ x, syn$syn)
se.syn  <- summary(fit.syn)$coefficients[, "Std. Error"]
se.corr <- calibrated.se(se.syn, syn$sigma.beta.raw)

data.frame(term    = names(se.syn),
           se.syn  = round(se.syn, 4),
           se.corr = round(se.corr, 4),
           noise   = round(unname(syn$sigma.beta.raw), 4))
```

`fidelity.inferential()` does this for you, and will **error** rather than quietly
return an uncorrected interval if the object cannot name its own noise scale:

```{r}
m <- fidelity.inferential(syn, orig = d, formula = y ~ x, calibrate = TRUE)
round(rbind(naive = m$se.syn, corrected = m$se.corr), 4)
```

## 3. How standard error grows as the privacy budget decreases

As epsilon falls the noise grows, and so does the corrected standard error. The
naive one barely moves, which is the problem: it is blind to the
mechanism.

```{r}
v.eps <- c(1, 2, 5, 10)
out <- t(vapply(v.eps, function(eps) {
  s   <- gen.syn.dp.ols.public(d, y ~ x, epsilon = eps, delta = 1e-6,
                               B.y = 8, seed = 1)
  fit <- lm(y ~ x, s$syn)
  se  <- summary(fit)$coefficients["x", "Std. Error"]
  c(eps     = eps,
    se.syn  = se,
    se.corr = calibrated.se(se, s$sigma.beta.raw[["x"]]))
}, numeric(3)))
round(out, 4)
```

## 4. Empirical coverage of confidence intervals

The point of the correction is that the interval should cover the truth at its
nominal rate. Generate data with a known coefficient, release it under DP, refit,
and check.

```{r}
n <- 400
beta.true <- 0.7
n.reps <- 50   # small for vignette build time; scale up offline

cover <- function(eps) {
  res <- vapply(seq_len(n.reps), function(k) {
    set.seed(100 + k)
    dk <- data.frame(x = rnorm(n))
    dk$y <- beta.true * dk$x + rnorm(n, sd = 0.5)
    s   <- gen.syn.dp.ols.public(dk, y ~ x, epsilon = eps, delta = 1e-6,
                                 B.y = 6, seed = k)
    fit  <- lm(y ~ x, s$syn)
    est  <- coef(fit)[["x"]]
    se.s <- summary(fit)$coefficients["x", "Std. Error"]
    se.c <- calibrated.se(se.s, s$sigma.beta.raw[["x"]])
    c(naive = abs(est - beta.true) < 1.96 * se.s,
      corr  = abs(est - beta.true) < 1.96 * se.c)
  }, numeric(2))
  rowMeans(res)
}

cover(10)
```

The corrected intervals reach approximately the nominal 95%. The naive ones undercover,
because they omit the DP noise variance entirely.

## 5. Scope and applicability

- The correction is exact for the routes whose refit recovers `beta.dp` exactly:
  `gen.syn.dp.ols.public()`, `gen.syn.dp.ridge.public()` and the Gaussian
  `gen.syn.dp.projected.subagg()`.
- **Route 3 bypasses this function entirely.**
  `gen.syn.dp.suffstats()` returns its own inference pair, `$beta.infer` and
  `$Cov.beta.infer`, the interval being centred at the unpenalised solve of the
  released statistics rather than at the ridge point estimate. Add
  `Cov.beta.infer[j, j]` to the synthetic variance directly, as that help page
  sets out. Centring on `beta.dp` instead is what under-covered at the
  strong-shrinkage budgets before the recentring fix.
- It corrects the **coefficient** channel only. A noisy dispersion widens
  `se.syn` on its own, and that is already in the naive term.
- `gen.syn.dp.projected()` (V1) carries **no formal DP guarantee**, so calibrating
  its intervals gives you calibrated uncertainty about a release with no privacy
  claim. It is a baseline rather than a mechanism. See `vignette("dp-guide")`.

## 6. See also

- `vignette("dp-guide")` for choosing a DP route.
- `vignette("getting-started")` for the non-DP projection.
