---
title: "Choosing a DP route from the assumptions you can defend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a DP route from the assumptions you can defend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(synthexpln)
set.seed(1)
```

## 1. The only question that matters

Differential privacy calibrates noise to a **global sensitivity**: the largest
the released quantity can move across *any* pair of neighbouring datasets. That
is a statement about a universe of possible datasets, so it can only come from
something you fixed **before you saw the data**.

Every DP route in this package therefore begins by asking you to name that
something. There are exactly three answers, and choosing among them is the whole
of the design work.

| You can assume | Route | Needs |
|---|---|---|
| The design is public **and** the response has a public bound | **Route 1**, exact and least noisy | `x.public = TRUE`, `B.y` |
| Neither, or only a public box on the coefficients | **Route 2**, subsample-and-aggregate | `clip.lo`, `clip.hi`, a `dispersion` |
| The design is **sensitive** but the response admits a public clip | **Route 3**, whitened sufficient statistics | `route = "suffstat"`, `W`, `B.x`, `y.centre`, `y.scale`, `y.clip` |

Route 1 is exact and least noisy wherever a genuine `B.y` exists. Route 2 is the
only route on an unbounded response, since it needs no bound on `|y|` at all.
Route 3 keeps the design private. Its standing against Route 2 depends on the
budget and on the width of the public coefficient box, and it is ahead of the
central box at the smallest budget and at every budget on a structurally
bounded response. Section 5 below covers the remaining case, a sensitive
design released under DP alongside the coefficient.

Routes 1 and 2 are selected by the shape of the problem, from what you supply.
Route 3 is **asked for by name**, `route = "suffstat"`, and is not inferred
from its constants. Its lead over Route 2 reverses with the budget, ahead at
small epsilon and behind from about `epsilon = 5` where the public bound is a
clip on an unbounded-support response, so a
wrapper that chose it for you would be picking a privacy-utility trade-off that
is yours to make.

If you can supply none of them, you do not have a mechanism, and `dp.project()`
will tell you so rather than inventing one.

```{r error = TRUE}
d <- data.frame(y = rnorm(500), x = rnorm(500))
dp.project(d, y ~ x, epsilon = 10)
```

That refusal is the most important line in the package. A bound read off the
sample (`B.y <- 1.5 * max(abs(y))`, say) makes the noise scale a function of the
very data it protects. Neighbouring datasets then get different noise, no
sensitivity bound holds across the pair, and there is **no theorem to
instantiate**. The release is not weakly private. It is not private at any
epsilon, and a smaller epsilon does not help.

## 2. Route 1: a public design and a bounded response

Where the covariates are public and the response has a genuine bound from the
domain, ordinary least squares has a clean global sensitivity. Replacing one
row's response moves `X'y` by `x_i (y_i' - y_i)` and nothing else, so

$$\Delta_\beta = 2 B_y \max_i \|(X'X)^{-1} x_i\|_2,$$

a function of the public design and the public bound alone. This route is exact
and it is the least noisy available, so prefer it whenever `B.y` is real.

`B.y` must come from the domain: a physiological range, a reporting cap, a
structurally bounded instrument. Below, `y` is standard normal *by construction*,
so six standard deviations bound it. That is a fact about the simulation on the
line above, not a quantity read off `d$y`.

```{r}
syn <- gen.syn.dp.ols.public(d, y ~ x, epsilon = 10, delta = 1e-6,
                             B.y = 6, seed = 1)

# the analyst's refit recovers the DP coefficient EXACTLY
all.equal(unname(coef(lm(y ~ x, syn$syn))), unname(syn$beta.dp))

# and the DP dispersion too
all.equal(summary(lm(y ~ x, syn$syn))$sigma^2, syn$sigma2.dp)
```

`gen.syn.dp.ridge.public()` is the same route with `A = X'X + lambda I` in place
of `X'X`, for ill-conditioned designs. Its sensitivity reduces to the OLS one as
`lambda` goes to zero.

### When Route 1 does not exist at all

If the response is genuinely unbounded, **no finite `B.y` exists**. Any Gaussian
outcome is unbounded. Route 1 is then not merely noisier, it is undefined, and
the accurate statement is to say so and use Route 2. Where you do want Route 1 on an
unbounded response, you must clip `y` to a public range as an explicit
preprocessing step, which moves the estimand to clipped-OLS and must be reported
as such.

## 3. Route 2: subsample-and-aggregate

Route 2 needs **no bound on the response at all**. Rows are partitioned into `m`
blocks by position under a fixed public seed, the model is fit per block, and
each coordinate is aggregated by a Widened Winsorized Mean over a **public box**.

One row changes exactly one block. That block's coefficient is clipped into the
box, so it moves by at most `hi_j - lo_j`, and the mean over `m` blocks moves by
at most `s_j = (hi_j - lo_j) / m`. No design constant and no statistic of the
data appears anywhere in that argument, so it holds under **full-row**
replacement: whether the neighbour changed `x`, `y`, or both, and however large
the response is.

That is the reason the mechanism exists (Nissim, Raskhodnikova and Smith
2007 designed it for statistics with no bounded global sensitivity).

The cost is that you must supply the box. That cost is unavoidable and deliberate,
since the old API needed no bounds at all, which is how the defect remained unnoticed.

```{r}
n  <- 1000
dg <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
dg$y <- 1 + 0.5 * dg$x1 - 0.3 * dg$x2 + rnorm(n, 0, 2)

syn2 <- gen.syn.dp.projected.subagg(
  dg, y ~ x1 + x2, epsilon = 10, family = gaussian(),
  clip.lo = -3, clip.hi = 3,            # PUBLIC box on the coefficients
  dispersion = dp.dispersion.public(
    4, source = "the instrument's quoted variance, published in the manual"),
  seed = 1)

all.equal(unname(coef(lm(y ~ x1 + x2, syn2$syn))), unname(syn2$beta.dp))
```

Route 2 publishes the covariates **as they stand**, so it does assume the design
is public. That is a statement about the *output*, not about the calibration, and
the two must not be conflated. Where the covariates are themselves sensitive, see
section 5.

## 4. The dispersion is a parameter, and DP must cover it

A GLM has two parameters, not one. Until version 0.2.0 only the coefficient passed
through a DP channel here and the dispersion rode into the release on the real
response, so the dispersion must now be **declared**, and the two defensible answers
carry different privacy claims.

```{r error = TRUE}
gen.syn.dp.projected.subagg(dg, y ~ x1 + x2, epsilon = 10,
                            family = gaussian(), clip.lo = -3, clip.hi = 3)
```

**Public** (no budget). The value is external knowledge, or the owner has *already
published* it: journal articles routinely report an $R^2$ or a residual SD
alongside a coefficient table. Post-processing permits any function of the DP
output and public information, so re-using it spends no budget and discloses
nothing new. Where the owner published it themselves, the claim is mu-GDP
*conditional* on that publication, and a paper must say so.

**Private** (costs). Nothing has been published, so it is paid for: `sigma^2`
rides along as a `(p+1)`-th aggregated coordinate over its own public box. The L2
sensitivity rises from `sqrt(p)` to `sqrt(p+1)`, so the coefficient noise rises
by about 8% at `p = 6`.

```{r}
pub <- dp.dispersion.public(4, source = "instrument variance, from the manual")
prv <- dp.dispersion.private(0, 25, source = "the instrument cannot resolve
                                              below an SD of 5, so sigma^2 <= 25")
pub
prv
```

`source` is **required** and is printed. That is the load-bearing part of the
design, not decoration. The unsound value still exists as a construction, but to
use it you now have to write

```r
dp.dispersion.public(var(residuals(fit)), source = "the sample")
```

which is self-indicting on the page, and which any reader of the code will catch.
A design that makes the defect unwriteable without saying so prevents the whole
class, where a single fix closes only the instance found.

## 5. Sensitive covariates

When the covariates must themselves be protected, `gen.syn.dp.full()` (Gaussian)
and `gen.syn.dp.logistic()` (binary) release them under DP and compose that with
the coefficient release. Every channel is a Gaussian mechanism, so the whole
release composes in **one** mu-GDP quadrature:

$$\mu^2 = \mu_\text{marg}^2 + \mu_\text{corr}^2 + \mu_\text{coef}^2.$$

```{r}
d4 <- data.frame(y = rnorm(600),
                 age = rnorm(600, 50, 10),
                 income = rnorm(600, 100, 20))
x.spec <- list(age    = list(type = "continuous", bounds = c(0, 100)),
               income = list(type = "continuous", bounds = c(40, 160)))

syn4 <- gen.syn.dp.full(d4, y ~ age + income, x.spec = x.spec,
                        epsilon = 10, delta = 1e-6,
                        clip.lo = -1, clip.hi = 1,    # PUBLIC coefficient box
                        disp.lo = 0,  disp.hi = 4,    # PUBLIC sigma^2 box
                        seed = 1)

syn4$mu.channels             # the three channels
syn4$mu.check                # recomposed in quadrature
syn4$mu.total                # == mu.check, exactly
```

Note these variants need **no `B.y`**. Subsample-and-aggregate takes its
sensitivity from the coefficient box, so a bound on the response buys nothing
here.

### When the response does admit a public clip: Route 3

`gen.syn.dp.full()` above releases the covariates themselves. Route 3 takes the
other option on the same problem, keeping the design private and releasing the
whitened sufficient statistics instead. It needs a public whitening matrix `W`,
a public bound `B.x` on `||W x||` over the support box, and a public clip on the
response. Ask for it by name:

```{r}
d5 <- data.frame(x1 = rnorm(400), x2 = rnorm(400))
d5$y <- 1 + 0.5 * d5$x1 - 0.3 * d5$x2 + rnorm(400)

W <- diag(3)
dimnames(W) <- list(c("(Intercept)", "x1", "x2"),
                    c("(Intercept)", "x1", "x2"))

syn5 <- dp.project(d5, y ~ x1 + x2, epsilon = 10, route = "suffstat",
                   W = W, B.x = sqrt(1 + 2 * 6^2),   # six sd on each standard normal
                   y.centre = 0, y.scale = 1, y.clip = c(-8, 8),
                   seed = 7)

syn5$variant                 # V6
syn5$mu.channels             # vech, X'y, y'y, and the eigenvalue channel
syn5$beta.infer              # the recentred inference pair
```

Every one of `W`, `B.x`, and `y.clip` must be public, on the same footing as
`B.y` in Route 1. They are forwarded rather than defaulted, so omitting one
reaches Route 3's own refusal naming what is missing. For the tuning knobs
beyond these constants (`mu.split`, `lambda`, `eig.split`), call
`gen.syn.dp.suffstats()` directly.

`mu.split` moves budget between the channels. The default starves the covariate
channels in favour of the coefficient, because the analyst's refit recovers
`beta_DP` to double-precision machine epsilon and so the joint law of the synthetic `X` does
not touch inferential fidelity:

```{r}
# default: inferential fidelity
c(marg = 0.05, corr = 0.05, coef = 0.90)

# shift it when the release must also describe X well
descriptive <- gen.syn.dp.full(d4, y ~ age + income, x.spec = x.spec,
                               epsilon = 10, delta = 1e-6,
                               mu.split = c(marg = 0.25, corr = 0.25, coef = 0.50),
                               clip.lo = -1, clip.hi = 1,
                               disp.lo = 0, disp.hi = 4, seed = 1)
descriptive$mu.channels
```

The split moves budget, not cost: `mu.total` is identical either way.

## 6. Why mu, and not epsilon

Every channel is calibrated in **mu-Gaussian DP** (Dong, Roth and Su 2022) and
the `(epsilon, delta)` label is read off the exact dual at the very end. Two
reasons.

**Composition.** Gaussian DP composes in quadrature. The classical
`(epsilon, delta)` pair does not: two mechanisms give `(2 eps, 2 delta)` by basic
composition, a bound that degrades linearly and that no choice of mechanism
recovers from. Under mu, `delta` no longer accumulates across channels at all.

**The classical multiplier is only valid below `eps = 1`.** The Dwork-Roth
Theorem A.1 bound `sqrt(2 log(1.25/delta))/eps` silently *under-noises* above
that. At `eps = 10, delta = 1e-6` it certifies only `delta = 1.9e-6`, nearly
twice what was asked for:

```{r}
eps <- 10; delta <- 1e-6
mu.classical <- 1 / (sqrt(2 * log(1.25 / delta)) / eps)
synthexpln:::.gdp.delta(eps, mu.classical)     # what it ACTUALLY certifies

mu.exact <- synthexpln:::.gdp.mu.from.eps.delta(eps, delta)
synthexpln:::.gdp.delta(eps, mu.exact)         # what was asked for
```

## 7. What is NOT differentially private

`gen.syn.dp.projected()` (V1) calibrates its noise to the DFBETA **local**
sensitivity, measured on the realised data. It carries **no formal DP guarantee**
at the stated epsilon or at any other, it warns on every call, and `dp.project()`
does not dispatch to it. It is retained as an *empirical-DP baseline*, which is a
real and useful quantity to compare against, and nothing more.

`sensitivity.local()` likewise computes a real quantity. It measures how much one
row of *this* dataset moves the fit, which is a sound influence diagnostic and the
natural yardstick for an empirical-versus-worst-case comparison. **Do not calibrate
a release to it.**

The same applies to `sens.method = "local"` on `gen.syn.dp.full()` and
`gen.syn.dp.logistic()`, kept only so that pre-0.2.0 runs can be reproduced. It
warns on every call.

## 8. Auditing the claim

`dp.audit()` runs a canary distinguishing game on a neighbouring pair. For a
mu-GDP mechanism the optimal distinguisher's AUC is at most
$\Phi(\mu/\sqrt{2})$, which is a far smaller target than the accuracy bound
usually quoted and can actually detect a leak.

Watch for an attack advantage that is **flat in epsilon**. That is the
fingerprint of a channel the noise did not reach, and it is how the
virtual-response leak was caught: the old Route 2 release scored AUC 1.000 at
*every* epsilon, because the real response was being carried into the release
untouched and epsilon was disconnected from it.

## 9. See also

- `vignette("getting-started")` for the non-DP projection.
- `vignette("calibrated-inference")` for correcting standard errors under DP.
- Li (2026) "Conditional moment conditions to reproduce statistical inference
  from synthetic data", the accompanying paper, `vignette("synthexpln-paper")`.
