---
title: "Getting started with synthexpln"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with synthexpln}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(synthexpln)
set.seed(1)
```

## 1. The problem

Analysts often release regression results computed on sensitive data that cannot be shared directly.
That raises the constructive question of whether synthetic data can preserve the inferential
properties of the original analysis, so that others may verify and build upon the reported findings.

**Projection-based synthetic data** offers a principled solution. The key insight, formalised in
the sufficient-statistics projection theorem of Li (2026), is that if we condition synthetic data
on the regression coefficients estimated from the original data, we can recover those coefficients
when fitting the same model to the synthetic data, to IEEE 754 double-precision machine epsilon
(about 2.2e-16) under a Gaussian identity link and to Newton tolerance for other GLM families. The synthetic data is then
**inferentially faithful**, preserving the analyst's point estimates. Distributional
realism is a separate property that exact coefficient recovery does not imply, and the paper's
central result is that optimising for it does not deliver inferential fidelity.

When combined with differential privacy, the analyst can add noise to the coefficients and
release a differentially private version, while still allowing synthetic data to be generated
that matches the published (noisy) coefficients to the same precision.

## 2. Non-DP projection

Consider a simple regression example. Suppose the original analyst has 500 observations
with a response and three covariates:

```{r}
n <- 500
d <- data.frame(
  y  = NA,
  x1 = rnorm(n),
  x2 = rbinom(n, 1, 0.4),
  x3 = runif(n)
)
d$y <- 1 + 0.5 * d$x1 - 0.3 * d$x2 + 0.2 * d$x3 + rnorm(n, 0.5)
```

First, the analyst fits a regression model and records the coefficients:

```{r}
fit.orig <- lm(y ~ x1 + x2 + x3, d)
coef(fit.orig)
```

Now we use `projection()` to generate synthetic data conditioned on those coefficients.
The function automatically extracts the model formula and fits coefficients, then projects
synthetic data onto the constraint surface:

```{r}
syn <- projection(d, y ~ x1 + x2 + x3)
coef(lm(y ~ x1 + x2 + x3, syn$syn))
```

Notice that the refitted coefficients match the originals to within a few units in the last place
of IEEE 754 double precision. Measured across 4,800 Gaussian runs spanning sample sizes from 100 to
5000, 2 to 8 predictors, and independent through strongly correlated designs, the deviation relative
to the size of the coefficient vector had a median of 3.9e-16 and a worst case of 3.1e-15, which is
13.9 ulp. That is the Gaussian identity-link case; other families are looser by orders of magnitude,
and `?projection` tabulates them.
This is the core guarantee: the synthetic data preserves the published coefficients.

## 3. Differentially private release

In practice, the analyst may want to add privacy protection before release.
`dp.project()` wraps `projection()` and adds differentially private noise to the
coefficients.

It will not, however, guess a privacy assumption on your behalf. Every sound DP
route rests on something fixed **before the data are seen**, and a wrapper cannot
invent one, so a bare call is refused:

```{r error = TRUE}
dp.project(d, y ~ x1 + x2 + x3, epsilon = 10)
```

Supply the assumption you can actually defend. Here the covariates are public and
`y` is bounded by construction, so `B.y` is a fact about the simulation rather
than a quantity read off `d$y`:

```{r}
syn.dp <- dp.project(d, y ~ x1 + x2 + x3,
                     epsilon = 10, delta = 1e-6,
                     x.public = TRUE, B.y = 20, seed = 1)
syn.dp$variant
coef(lm(y ~ x1 + x2 + x3, syn.dp$syn))
syn.dp$beta.dp
```

The returned object contains:

- `syn$syn`: the synthetic data frame, which is the DP release
- `beta.dp`: the differentially private coefficients
- `sigma.beta.raw`: the per-coefficient noise sd, for inference calibration
- `mu.total`, `mu.check`: the mu-GDP accounting

When others refit the model on this synthetic data, they recover `beta.dp`
exactly, the noisy coefficients the analyst intended to publish. Note that
`beta.hat` is also returned but is the **non-private** full-data fit, carried for
evaluation only: it must not be composed downstream.

See `vignette("dp-guide")` for choosing among the three sound routes.

## 4. Calibrated standard errors

The standard errors computed from the synthetic data understate the true
uncertainty, because they know nothing about the privacy noise that produced the
coefficients in the first place. Correct them with `calibrated.se()`, passing the
per-coefficient noise sd the object reports:

```{r}
fit.syn <- lm(y ~ x1 + x2 + x3, syn.dp$syn)
se.syn  <- summary(fit.syn)$coefficients[, "Std. Error"]
se.corr <- calibrated.se(se.syn, syn.dp$sigma.beta.raw)
data.frame(term = names(se.syn), se.syn, se.corr)
```

`$sigma.beta.raw` is the noise sd on the **raw** scale of each coefficient. Pass
that and nothing else: there is no column scale to divide by, and dividing by one
would shrink the correction and understate the uncertainty of a private release.
See `vignette("calibrated-inference")`.

The corrected standard errors are larger, reflecting the uncertainty the privacy
mechanism actually introduced. These are what to report.

## 5. Auditing the DP guarantee

To check that the empirical privacy guarantee matches the theoretical one,
use `dp.audit()`. This function is computationally expensive and is gated behind the
`SYNTHEXPLN_FULL_TESTS` environment variable. See `vignette("dp-guide")` for details.

## 6. Next steps

- Read `vignette("dp-guide")` for a deeper dive into privacy tuning and auditing.
- Read `vignette("calibrated-inference")` for guidance on valid statistical inference
  with calibrated standard errors.
- See Li (2026) for the theoretical foundations.
